---
title: 'Week 2 Lab: Introduction to Data'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages


install.packages("statar")
install.packages("ggplot2")
install.packages("dplyr")

```{r}
library(statar)
library(ggplot2)
library(dplyr)
library(statsr)
```


```{r}
?nycflights
```

### Data

Airline on-time data for all flights departing NYC in 2013. Also includes useful 'metadata' on airlines, airports, weather, and planes.

```{r}
data("nycflights")
```

```{r}
str(nycflights)
```

## Analysis

### Departure delays in flights to Raleigh-Durham (RDU)
```{r hist-dep-delay}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram()
```

```{r hist-dep-delay-bins-15}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 15)

```

```{r hist-dep-delay-bins-150}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 150)
```

Binwidth was required since the histogram needs to be zoomed in more to observe how much space before 0 in the x-axis.

### Create another vector to filter dest has valus Raleigh-Durham International Airport (RDU)

```{r create-rdu_flights-filter-RDU}
rdu_flights <- nycflights %>% 
  filter(dest == "RDU")
ggplot(data = rdu_flights, aes(x = dep_delay)) +
  geom_histogram()
```

```{r rdu_flights-Sum}
rdu_flights %>% 
  summarise(mean_dd = mean(dep_delay), sd_dd = sd(dep_delay), n = n(), var(dep_delay), IQR(dep_delay), median(dep_delay), min(dep_delay), max(dep_delay))

```

### Create dataframe for SFO_FEB

```{r create-dataframe-sfo_feb_flights}
sfo_feb_flights <- nycflights %>% 
  filter(dest == "SFO", month == 2)
```

```{r sfo_feb_flights}
dim(sfo_feb_flights)
```

How many flights did head to SFO in February ?
68

```{r hist-sfo_feb_flighs-bins-10}
ggplot(data = sfo_feb_flights, aes(x = arr_delay)) +
  geom_histogram(binwidth = 10)

```

```{r summary-sfo_feb_flights}
sfo_feb_flights %>% 
  summarise(mean(arr_delay), median(arr_delay), max(arr_delay))
```
Some flights delayed more than 2 hours 

```{r summary-custom-list-origin}
rdu_flights %>% 
  group_by(origin) %>% 
  summarise(mean_dd = mean(dep_delay), sd_dd = sd(dep_delay), n = n())
```

```{r calculate the median arr_delay sfo_feb_flights}
sfo_feb_flights %>% 
  group_by(carrier) %>% 
  summarise(med_ad = median(arr_delay), iqr_ad = IQR(arr_delay)) %>% 
  arrange(desc(iqr_ad))


```
Delta and United airlines has the highest IQR of arrival delays

### Departure delays over months

stakeholders ask Which month has the highest average delay departing from an NYC airport?

```{r calculate monthly mean averages for departure delays}
nycflights %>% 
  group_by(month) %>% 
  summarise(mean_dd = mean(dep_delay), iqr_dd = IQR(dep_delay)) %>% 
  arrange(desc(mean_dd))

```
July has the highest average departure delay from an NYC airport

stakeholder ask which month has the highest median departure delay from an NYC airport ?

```{r calculate monthly median for departure delays }
nycflights %>% 
  group_by(month) %>% 
  summarise(med_dd = median(dep_delay)) %>% 
  arrange(desc(med_dd))

```
December has the highest median departure delay from an NYC airport

stakeholder ask which is more reliable masure for deciding which month(s) to avoid flying if customs really dislike delayed flights (mean or median), and why?

```{r delay month box}
ggplot(nycflights, aes(x = factor(month), y = dep_delay)) +
  geom_boxplot()
```

Mean would be more reliable as it gives us the true average

### On time departure rate for NYC airports
which NYC airport would customers choose to fly out of if they were selecting an airport simply based on time departure percentage?

```{r dep_type}
nycflights <- nycflights %>% 
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))

```

```{r}
nycflights %>% 
  group_by(origin) %>% 
  summarise(ot_dep_rate = sum(dep_type == "on time") / n()) %>% 
  arrange(desc(ot_dep_rate))
```

```{r}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar()
```

they were choose LGA 

```{r fastest-avg-speed-tailnum}
nycflights <- nycflights %>% 
  mutate(avg_speed = distance / (air_time/60))
nycflights %>% 
  select(avg_speed, tailnum) %>% 
  arrange(desc(avg_speed))

```

```{r avg-speed-dist-scatter}
ggplot(nycflights, aes(x = distance, y = avg_speed)) +
  geom_point()
```

There is an overall postive association between distance and average speed. 

```{r on-time-arr-perc}
nycflights %>% 
  mutate(arr_type = ifelse(arr_delay <= 0, "on time", "delayed")) %>% 
  mutate(dep_type = ifelse(dep_delay <= 0, "on time", "delayed")) %>% 
  select(arr_type,dep_type) %>% 
  table()

```

3508/ (3508 + 9291) = 0.2740839
