---
title: 'Week 2 Lab: Introduction to Data'
author: 'Your Name'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
# Set global knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Set seed for reproducibility
set.seed(123)
```

# Setup

## Package Management
```{r packages}
# Function to install and load required packages
load_packages <- function(packages) {
  new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(new_packages)) install.packages(new_packages)
  invisible(lapply(packages, library, character.only = TRUE))
}

# Required packages
required_packages <- c("statar", "ggplot2", "dplyr", "statsr", "tidyverse")
load_packages(required_packages)
```

# Data Loading and Initial Exploration

```{r load_data}
# Load the nycflights dataset
data("nycflights")

# Display structure of the data
glimpse(nycflights)
```

# Data Cleaning and Preparation

```{r data_preparation}
# Create clean version of flights data with derived variables
clean_flights <- nycflights %>%
  # Handle missing values
  drop_na() %>%
  # Create new variables
  mutate(
    dep_type = if_else(dep_delay < 5, "on time", "delayed"),
    arr_type = if_else(arr_delay <= 0, "on time", "delayed"),
    avg_speed = distance / (air_time/60),
    month = factor(month, levels = 1:12, 
                  labels = month.abb[1:12])  # Convert month to factor with labels
  )
```

# Analysis Functions

```{r analysis_functions}
#' Summarize flight delays
#' @param data DataFrame containing flight data
#' @param delay_col Name of the delay column to analyze
#' @return DataFrame with summary statistics
summarize_delays <- function(data, delay_col) {
  data %>%
    summarise(
      n_flights = n(),
      mean_delay = mean({{delay_col}}, na.rm = TRUE),
      median_delay = median({{delay_col}}, na.rm = TRUE),
      sd_delay = sd({{delay_col}}, na.rm = TRUE),
      iqr_delay = IQR({{delay_col}}, na.rm = TRUE),
      min_delay = min({{delay_col}}, na.rm = TRUE),
      max_delay = max({{delay_col}}, na.rm = TRUE)
    ) %>%
    round(2)
}

#' Create standardized delay distribution plot
#' @param data DataFrame containing flight data
#' @param x_var Name of the variable to plot
#' @param title Plot title
#' @param binwidth Bin width for histogram
create_delay_plot <- function(data, x_var, title, binwidth = 15) {
  ggplot(data, aes(x = {{x_var}})) +
    geom_histogram(
      binwidth = binwidth,
      fill = "steelblue",
      color = "white"
    ) +
    labs(
      title = title,
      x = "Delay (minutes)",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10)
    )
}
```

# Analysis

## Departure Delays Analysis

### Overall Departure Delay Distribution
```{r overall_delays}
create_delay_plot(
  clean_flights, 
  dep_delay,
  "Distribution of All NYC Flight Departure Delays"
)
```

### RDU Flights Analysis
```{r rdu_analysis}
# Filter flights to RDU
rdu_flights <- clean_flights %>% 
  filter(dest == "RDU")

# Create visualization
create_delay_plot(
  rdu_flights,
  dep_delay,
  "Distribution of Departure Delays for Flights to RDU"
)

# Calculate summary statistics
rdu_summary <- rdu_flights %>%
  summarize_delays(dep_delay)

# Display summary
knitr::kable(rdu_summary, caption = "Summary Statistics for RDU Flights")
```

### SFO February Flights Analysis
```{r sfo_feb_analysis}
# Filter February flights to SFO
sfo_feb_flights <- clean_flights %>% 
  filter(dest == "SFO", month == "Feb")

# Analyze arrival delays by carrier
sfo_carrier_delays <- sfo_feb_flights %>%
  group_by(carrier) %>%
  summarise(
    median_delay = median(arr_delay, na.rm = TRUE),
    iqr_delay = IQR(arr_delay, na.rm = TRUE)
  ) %>%
  arrange(desc(iqr_delay))

# Create visualization
ggplot(sfo_feb_flights, aes(x = carrier, y = arr_delay)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Arrival Delays by Carrier for February SFO Flights",
    x = "Carrier",
    y = "Arrival Delay (minutes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Monthly Delay Patterns

```{r monthly_patterns}
# Calculate monthly delay statistics
monthly_delays <- clean_flights %>%
  group_by(month) %>%
  summarise(
    mean_delay = mean(dep_delay, na.rm = TRUE),
    median_delay = median(dep_delay, na.rm = TRUE),
    n_flights = n()
  )

# Create monthly delay visualization
ggplot(clean_flights, aes(x = month, y = dep_delay)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Monthly Distribution of Departure Delays",
    x = "Month",
    y = "Departure Delay (minutes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Airport Performance Analysis

```{r airport_performance}
# Calculate on-time departure rates by origin airport
airport_performance <- clean_flights %>%
  group_by(origin) %>%
  summarise(
    n_flights = n(),
    on_time_rate = mean(dep_type == "on time", na.rm = TRUE),
    avg_delay = mean(dep_delay, na.rm = TRUE)
  ) %>%
  arrange(desc(on_time_rate))

# Create airport performance visualization
ggplot(clean_flights, aes(x = origin, fill = dep_type)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "On-Time Performance by NYC Airport",
    x = "Airport",
    y = "Proportion of Flights",
    fill = "Departure Status"
  ) +
  theme_minimal()
```

## Speed and Distance Analysis

```{r speed_distance}
# Create speed vs distance scatter plot
ggplot(clean_flights, aes(x = distance, y = avg_speed)) +
  geom_point(alpha = 0.1, color = "blue") +
  geom_smooth(method = "loess", color = "red") +
  labs(
    title = "Relationship Between Flight Distance and Average Speed",
    x = "Distance (miles)",
    y = "Average Speed (mph)"
  ) +
  theme_minimal()

# Find fastest flights
top_speeds <- clean_flights %>%
  select(tailnum, avg_speed, distance) %>%
  arrange(desc(avg_speed)) %>%
  head(10)

knitr::kable(top_speeds, caption = "Top 10 Fastest Flights")
```

## Departure vs Arrival Delay Analysis

```{r dep_arr_analysis}
# Create contingency table
delay_table <- table(
  clean_flights$dep_type,
  clean_flights$arr_type,
  dnn = c("Departure", "Arrival")
)

# Display formatted table
knitr::kable(delay_table, caption = "Departure vs Arrival Delay Contingency Table")

# Calculate conditional probability
on_time_arr_given_delayed_dep <- delay_table["delayed", "on time"] / 
  sum(delay_table["delayed", ])

cat("Probability of on-time arrival given delayed departure:", 
    round(on_time_arr_given_delayed_dep, 4))
```
